### 2021-09-28 papers/thesis/resubmitted/ch1-introduction/introduction.tex
Fixed typos and (most) notes from AT corrections for intro

Univariate orthogonal polynomials (hereon also referred to as OPs) have been extensively involved in the development of multiple fields of computational and applied mathematics. For example, univariate OPs have been used to derive spectral methods to numerically solve one-dimensional differential equations (see e.g. \cite{trefethen2000spectral, canuto2007spectral, gottlieb1977numerical, boyd2001chebyshev, mason2002chebyshev, shen2011spectral, olver2013fast}). While there are many famous examples of univariate OPs -- such as the Jacobi polynomials, the Legendre polynomials and the Chebyshev polynomials but to name a few of the classical families \cite[\S18.3]{DLMF} -- the area of multivariate orthogonal polynomials has a smaller array of research.

One could say this is surprising, given that there is a long history of around 150 years associated with multivariate OPs, beginning with Hermite first presenting the multivariate Hermite polynomials in 1865 \cite{appel1926fonctions, ismail2017review}. Zernike polynomials \cite{zernike1934diffraction} were first introduced in 1934, as another example, that are a group of bivariate polynomials orthogonal on a unit circle. Koornwinder in 1975 described a method for constructing two-variable OPs from univariate OPs \cite{koornwinder1975two}. However, few books have been published over the years on the topic of multivariate orthogonal polynomials, with the notable exception of the book by Dunkl \& Xu \cite{dunkl2014orthogonal}. As Dumitriu, Edelman and Shuman said in 2007, \enquote{[multivariate orthogonal polynomials] are understudied, underapplied, and important applications may be being missed} \cite{dumitriu2007mops}, while in 2016, Olver opined that \enquote{there is still a long way to go before multivariate orthogonal polynomials can reach their full potential in applications}. He hypothesised that \enquote{[p]art of this neglect may be due to the fact that mutually orthogonal polynomials are not uniquely defined: there is no canonical ordering} and continued \enquote{[e]ven when explicit constructions are known, these are often unwieldy} \cite{olver2016review}.

@@

Spherical harmonics (SHs) are of course a long-established and famous group of functions defined on the surface of a sphere with certain useful properties (for example, they are orthogonal to each other on the unit sphere and form a basis for expanding functions defined on the sphere) and as a result are widely used in many scientific fields for solving PDEs including computer graphics (e.g. \cite{moon2008efficient, sloan2013efficient}), astrophysics (e.g. \cite{vasil2019tensor}), quantum theory (e.g. \cite{varshalovich1988quantum}), biochemistry (e.g.\cite{parimal2014application, basko1998application}), geosciences (e.g. \cite{fletcher2017data, hollerbach2013parity}) and meteorology (e.g. \cite{evans1998spherical, rubinstein2015scalar, wedi2013fast, ecmwf2020scalability, courtier1998ecmwf, silberman1954planetary, gottlieb1977numerical}). Spectral methods on the sphere involving spherical harmonics have been used for over 60 years \cite{silberman1954planetary}. Notably, spherical harmonics are also used as basis functions for the spectral transform method that makes up part of the model in the Integrated Forecasting System (IFS), which is used by ECMWF for their forecasts \cite{wedi2013fast}.

@@

The motivation for this project was to help address this problem while still utilising a spectral approach. More precisely, we aim to develop a sparse spectral method for solving PDEs on the spherical cap as a surface in 3D, with a simple extension to a spherical band. Together, these frameworks can be pieced together to create a spectral element method for the whole sphere, or further developed to investigate spectral methods on other spherical subdomains. This, however, is future work beyond the scope of this thesis. By spectral element method, we mean a finite element method (FEM) that uses high degree basis polynomials for its elements (this could also be referred to as a $p$-FEM with large $p$). In other words, we can use our spectral methods developed for the spherical band and cap elements as part of a finite element framework. By using this approach, one would be able to avoid having to complete the global spectral transform (in particular, the global Legendre transform) and instead be able to simply apply the local element transforms in parallel. Moreover, by still using a spectral approach, one can maintain the high accuracy and excellent error properties that such methods bring. Solving PDEs on a section of the sphere surface is still useful for numerical weather prediction in its own right too, where one could for example perform a more localised simulation near a pole while still using a \enquote{global} (in the domain sense) spectral method approach. Moreover, there are also prospective applications in physics, particularly in astrophysics, where solving PDEs on the sphere surface and working in spherical geometries is also desirable (e.g. \cite{vasil2019tensor, reinecke2013libsharp, beyer2014numerical, varshalovich1988quantum, slevinsky2018spectral, rubinstein2015scalar}).

Recently, a method for computing tensor fields in spherical coordinates using Jacobi polynomials has been proposed \cite{vasil2019tensor}. In this work the authors present the method for both the surface of the unit sphere and the three-dimensional generalisation of the unit ball. Their method involves using a spectral basis to represent functions too, choosing for the angular part of the basis to be the spin-weighted spherical harmonics (of which the spherical harmonics are a special case, with spin $0$) in spherical coordinates. By using these as their basis functions, they are able to derive sparse relations for the Laplacian operator, as well as for multiplication of the basis by $\cosphi$ and $\sinphi$ (where $\varphi$ is the polar angle from the $z$ axis in cartesian coordinates) which lead to operators for operations with angular dependencies involving these.

>>>>

Univariate orthogonal polynomials (hereon also referred to as OPs) have been extensively involved in the development of multiple fields of computational and applied mathematics. For example, univariate OPs have been used to derive spectral methods to numerically solve one-dimensional differential equations (see e.g. \cite{trefethen2000spectral, canuto2007spectral, gottlieb1977numerical, boyd2001chebyshev, mason2002chebyshev, shen2011spectral, olver2013fast}). While there are many famous examples of univariate OPs -- such as the Jacobi polynomials, with special cases of the Legendre polynomials and the Chebyshev polynomials but to name a few of the classical families \cite[\S18.3]{DLMF} -- the area of multivariate orthogonal polynomials has a smaller array of research.

One could say this is surprising, given that there is a long history of around 150 years associated with multivariate OPs, beginning with Hermite first presenting the multivariate Hermite polynomials in 1865 \cite{appel1926fonctions, ismail2017review}. Zernike polynomials \cite{zernike1934diffraction} were first introduced in 1934, as another example, that are a group of bivariate polynomials orthogonal on a unit circle. Koornwinder in 1975 described a method for constructing two-variable OPs from univariate OPs \cite{koornwinder1975two}. However, few books have been published over the years on the topic of multivariate orthogonal polynomials, with the notable exception of the book by Dunkl and Xu \cite{dunkl2014orthogonal}. As Dumitriu, Edelman and Shuman said in 2007, \enquote{[multivariate orthogonal polynomials] are understudied, underapplied, and important applications may be being missed} \cite{dumitriu2007mops}, while in 2016, Olver opined that \enquote{there is still a long way to go before multivariate orthogonal polynomials can reach their full potential in applications}. He hypothesised that \enquote{[p]art of this neglect may be due to the fact that mutually orthogonal polynomials are not uniquely defined: there is no canonical ordering} and continued \enquote{[e]ven when explicit constructions are known, these are often unwieldy} \cite{olver2016review}.

@@

Spherical harmonics (SHs) are of course a long-established and famous group of functions defined on the surface of a sphere with certain useful properties (for example, they are orthogonal to each other on the unit sphere, have sparse recurrences, and are eigenfunctions of the spherical Laplacian) and as a result are widely used in many scientific fields for solving PDEs including computer graphics (e.g. \cite{moon2008efficient, sloan2013efficient}), astrophysics (e.g. \cite{vasil2019tensor}), quantum theory (e.g. \cite{varshalovich1988quantum}), biochemistry (e.g.\cite{parimal2014application, basko1998application}), geosciences (e.g. \cite{fletcher2017data, hollerbach2013parity}) and meteorology (e.g. \cite{evans1998spherical, rubinstein2015scalar, wedi2013fast, ecmwf2020scalability, courtier1998ecmwf, silberman1954planetary, gottlieb1977numerical}). Spectral methods on the sphere involving spherical harmonics have been used for over 60 years \cite{silberman1954planetary}. Notably, spherical harmonics are also used as basis functions for the spectral transform method that makes up part of the model in the Integrated Forecasting System (IFS), which is used by ECMWF for their forecasts \cite{wedi2013fast}.

@@

The motivation for this project was to help address this problem while still utilising a spectral approach. More precisely, we aim to develop a sparse spectral method for solving PDEs on the spherical cap as a surface in 3D, with a simple extension to a spherical band. Together, these frameworks can be pieced together to create a spectral element method for the whole sphere, or further developed to investigate spectral methods on other spherical subdomains. This, however, is future work beyond the scope of this thesis. By spectral element method, we mean a finite element method (FEM) that uses high degree basis polynomials for its elements (this could also be referred to as a $p$-FEM with large $p$). In other words, we can use our spectral methods developed for the spherical band and cap elements as part of a finite element framework. By using this approach, one can avoid having to compute the global spectral transform (in particular, the global Legendre transform) and instead simply apply the local element transforms in parallel. Moreover, by still using a spectral approach, one can maintain the high accuracy and excellent error properties that such methods bring. Solving PDEs on a section of the sphere surface is still useful for numerical weather prediction in its own right too, where one could for example perform a more localised simulation near a pole while still using a \enquote{global} (in the domain sense) spectral method approach. Moreover, there are also prospective applications in physics, particularly in astrophysics, where solving PDEs on the sphere surface and working in spherical geometries is also desirable (e.g. \cite{vasil2019tensor, reinecke2013libsharp, beyer2014numerical, varshalovich1988quantum, slevinsky2018spectral, rubinstein2015scalar}).

Recently, a method for computing tensor fields in spherical coordinates using Jacobi polynomials has been proposed \cite{vasil2019tensor}. In this work the authors present the method for both the surface of the unit sphere and the three-dimensional generalisation of the unit ball. Their method involves using a spectral basis to represent functions too, choosing for the angular part of the basis to be the spin-weighted spherical harmonics (of which the spherical harmonics are a special case, with spin $0$) in spherical coordinates. By using these as their basis functions, they are able to derive sparse relations for the Laplacian operator, as well as for multiplication of the basis by $\cosphi$ and $\sinphi$ (where $\varphi$ is the polar angle from the $z$ axis in cartesian coordinates), which lead to operators for operations with angular dependencies involving these.
